A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-11 tex/codatadp.tex
More work on solution soundness, cleanup description of bisimilarity/cyclicity.

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Cyclic}
\(
\inferrule{
  t : \tau
  \quad
  \tau \in \Data
  \quad
  %\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset
  \Val \ec{t} = \mu x. t'
  \quad
  x \in FV( t' )
}{
  \bot
}
\)
\\[3.7ex]
\rn{Bisimilar}
\(
\inferrule{
 t_1, t_2 : \tau
 \quad
 \tau \in \Codata
 \quad
 %\tpath{\Ec}{\ec{t_1}} = \tpath{\Ec}{\ec{t_2}}\sigma \neq \emptyset
 \Val \ec{t_1} \vsim \Val \ec{t_2}
}{
 \Ec := \Ec, t_1 \teq t_2
}
\)
\end{tabular}
\caption{Rules for cyclicity and bisimilarity.
}
\label{fig:ab-rules}
\end{figure}

@@

Recall that the $\mu$-notation can be used to represent (possibly cyclic) terms and values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term is either of the form
$\mu x. \const{C}( \vec t )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$, 
then if $\vec v$ are well-type $\mu$-terms under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if it contains no free variables.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent modulo renaming of $\mu$-bound variables.
For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
where in the latter case, a $\mu$-term with a bound variable denotes an invalid term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms.
Informally, the terms mapped to by $\Val$ describe the (class of) values that terms in each equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$ of $\Ec$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$ for each equivalence class $\ec{x}$ of $\Ec$ 
(in other words, there are initially no constraints on the values for any equivalence class).
Then, $\Val$ is updated by applying the following rule exhaustively:

@@

Here, we write $FV( \Val )$ to denote the free variables occurring in $\Val$, which
excludes variables that occur beneath $\mu$-binders.
It is straightforward to show that the height of terms produced as a result of this rule
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

\begin{example}
Say that $\Ec$ contains (distinct) equivalence classes $\ec{x}$, $\ec{y}$, $\ec{z}$ and $\ec{w}$ where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$.
A possible sequence of applications of the above rule is shown below.
In the first column, we show the equivalence class that was considered,
and in the second column, we show the state of $\Val$.
\[\begin{array}{|l|@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l|}
\hline
\text{Eqc} & \multicolumn{3}{c}{\Val} & \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{y}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\hline
\end{array}\]
Informally, this indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( w, \const{C}( z, \ldots ))$ (resp. $\const{C}( z, \const{C}( w, \ldots ))$) for some value of $w$ and $z$.
$\Box$
\end{example}

Given $\Val$, the rules for cyclicity and bisimilarity (shown in Figure~\ref{fig:ab-rules}) can be stated as follows.
For cyclicity, if there is an equivalence class $\ec{t}$ of datatype type whose class of values $\Val \ec{t} = \mu x. t'$ are cyclic,
as indicated by the assumption $x \in FV( t' )$,
then we can conclude that $\Ec$ is unsatisfiable.
For bisimilarity, if there are two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
of co-datatype type whose classes of values $\Val \ec{t_1}$ and $\Val \ec{t_2}$ are equivalent modulo renaming of $\mu$-bound variables,
then we can conclude $t_1 \teq t_2$.
We demonstrate the bisimilarity rule concretely with the following example.

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned steps, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, v_{\ec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, v_{\ec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ v_{\ec{x}} \mapsto v_{\ec{\const{C}(y,x)}}, v_{\ec{\const{C}(y,x)}} \mapsto v_{\ec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ are observationally equivalent for any value of $y$.
\qed
\end{example}

@@

\begin{lemma}[Refutation Soundness] \label{lem:rs}
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Unify$_2$}, this is a consequence of distinctness.
For \rn{Cyclic}, this is a consequence of induction. \rem{elaborate?}
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Unify$_1$}, this is a consequence of injectivity.
For \rn{Bisimilar}, this is a consequence of co-induction. \rem{elaborate?}
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node $\Fc$,
then $\Ec$ is satisfiable $\thD$.

%and second show how to construct a valuation for the variables of $\Fc \cup \Gc$.

@@

%Given a set of disjoint paths $P$ where each $p \in P$
%we write $\interp{P}$ to denote the value

\rem{Revise : now $\Val$ is constructed by calculus.}

First, we require additional terminology concerning $\mu$-terms.
The \emph{interpretation of subvalue $u$ in $v$} is the value $v\interp{u}^\emptyset$ returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec t ) \in v$ indicates that $\mu x. C( \vec t )$ is the subvalue of $v$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
v\interp{x}^M & = & v\interp{\mu x. C( \vec t )}^M & x \not\in M, x \mapsto \mu x. C( \vec t ) \in v  \\
v\interp{x}^M & = & x & \text{otherwise} \\
v\interp{\mu x. C( \vec t )}^M & = & \mu x. C( v\interp{ \vec t }^{M \cup \{ x \}} ) & x \not\in M \\
v\interp{\mu x. C( \vec t )}^M & = & x & \text{otherwise}
\end{array}\]
Given a $\mu$-value $v$ of the form $\mu x. C( \vec w )$, 
a subvalue $u$ of $v$ is a \emph{bisimilar subvalue} if and only if $v\interp{u}^\emptyset \sim v$.
We say $v$ is \emph{normal} if and only if all of its bisimilar subvalues are $x$,
and all of its strict subvalues that do not contain $v$ are also normal.
For instance, the value $v$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is not normal, 
since 
$v\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( v\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( v\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( v\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$,
which is equivalent to $v$ modulo the renaming $\{ x \mapsto y, y \mapsto x \}$.
For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form for $v$})
that is observationally equivalent to $v$.
We construct $\nf{v}$ by replacing all of the bisimilar subvalues of $v$ with $x$,
and by replacing all of the non-normal closed subvalues of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.

@@

\(
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  c \not\sim c' \text{ for all } c' \in \vrange(\Val)
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto c \}
}
\)
Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.

We first argue that $\Val$ contains only to ground values (which include closed $\mu$-terms), in other words, $FV( \Val ) = \emptyset$.
First, since \rn{Split} does not apply to $\Fc$,
all equivalence classes of finite type contain constructor terms.
Thus, all variables in $FV( \Val )$ have infinite type.
%\rem{Require : all co-datatypes either have infinite cardinality, or have cardinality 1.}

>>>>

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Cyclic}
\(
\inferrule{
  t_1 : \tau
  \quad
  \tau \in \Data
  \quad
  %\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset
  \Val \ec{t_1} = \mu x. t_2
  \quad
  x \in FV( t_2 )
}{
  \bot
}
\)
\\[3.7ex]
\rn{Bisimilar}
\(
\inferrule{
 t_1, t_2 : \tau
 \quad
 \tau \in \Codata
 \quad
 %\tpath{\Ec}{\ec{t_1}} = \tpath{\Ec}{\ec{t_2}}\sigma \neq \emptyset
 \Val \ec{t_1} \vsim \Val \ec{t_2}
}{
 \Ec := \Ec, t_1 \teq t_2
}
\)
\end{tabular}
\caption{Rules for cyclicity and bisimilarity.
}
\label{fig:ab-rules}
\end{figure}

@@

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve terms containing $\mu$-bindings.
Recall that these bindings can be used for representing (possibly cyclic) values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure either of the form
$\mu x. \const{C}( \vec t )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
For consistency, we will also consider $\mu x. \const{c}$ to be a $\mu$-term
in the case where $c$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$
and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if it contains no free variables.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some renaming of $\mu$-bound variables.
%For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
either in the case when $\tau$ is a co-datatype or datatype type,
where in the latter case, such a $t$ containing a bound variable describes an infeasible class of values.
At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$ of $\Ec$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$ for each equivalence class $\ec{x}$ of $\Ec$ 
(in other words, there are initially no constraints on the values for any equivalence class).
Then, $\Val$ is updated by applying the following rule exhaustively:

@@

Here, we write $FV( \Val )$ to denote the \emph{free variables} occurring in $\Val$, which
excludes variables that occur beneath $\mu$-binders.
It is straightforward to see that the height of terms produced as a result of this rule
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

\begin{example}
Say that $\Ec$ contains (distinct) equivalence classes $\ec{x}$, $\ec{y}$, $\ec{z}$ and $\ec{w}$ where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$.
A possible sequence of applications of the above rule is shown below.
In the first column, we show the equivalence class that was considered,
and in the second column, we show the content of $\Val$.
\[\begin{array}{|l|@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l|}
\hline
\text{Eqc} & \multicolumn{3}{c}{\Val} & \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{y}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\hline
\end{array}\]
Informally, this indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( v_{\ec{w}}, \const{C}( v_{\ec{z}}, \ldots ))$ (resp. $\const{C}( v_{\ec{z}}, \const{C}( v_{\ec{w}}, \ldots ))$) 
for some value of $v_{\ec{w}}$ and $v_{\ec{z}}$.
$\Box$
\end{example}

Given $\Val$, the rules for cyclicity and bisimilarity (shown in Figure~\ref{fig:ab-rules}) can be stated as follows.
For cyclicity, if there is an equivalence class $\ec{t_1}$ of datatype type whose class of values $\Val \ec{t_1} = \mu x. t_2$ are cyclic,
as indicated by the assumption $x \in FV( t_2 )$,
then we can conclude that $\Ec$ is unsatisfiable.
For bisimilarity, if there are two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
of co-datatype type whose classes of values $\Val \ec{t_1}$ and $\Val \ec{t_2}$ are equivalent modulo renaming of $\mu$-bound variables,
then we can conclude $t_1 \teq t_2$.
We demonstrate the bisimilarity rule concretely with the following example.

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned steps, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, v_{\ec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu v_{\ec{\const{C}(y,x)}}. & \const{C}( v_{\ec{y}}, & \mu v_{\ec{x}}. & \const{C}( v_{\ec{y}}, v_{\ec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ v_{\ec{x}} \mapsto v_{\ec{\const{C}(y,x)}}, v_{\ec{\const{C}(y,x)}} \mapsto v_{\ec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ are observationally equivalent for any value of $v_{\ec{y}}$.
\qed
\end{example}

@@

\begin{lemma}[Refutation Soundness] \label{lem:rs}
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Unify$_2$}, this is a consequence of distinctness.
For \rn{Cyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Unify$_1$}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are observationally equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node $\Fc$,
then $\Ec$ is satisfiable $\thD$.
First, we require additional terminology concerning $\mu$-terms.
If $u$ is a subterm of $v$, then the \emph{interpretation of $u$ in $v$} is the term $v\interp{u}^\emptyset$ returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec t ) \in v$ indicates that $\mu x. C( \vec t )$ is the subterm of $v$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
v\interp{x}^M & = & v\interp{\mu x. C( \vec t )}^M & x \not\in M, x \mapsto \mu x. C( \vec t ) \in v  \\
v\interp{x}^M & = & x & \text{otherwise} \\
v\interp{\mu x. C( \vec t )}^M & = & \mu x. C( v\interp{ \vec t }^{M \cup \{ x \}} ) & x \not\in M \\
v\interp{\mu x. C( \vec t )}^M & = & x & \text{otherwise}
\end{array}\]
Given a $\mu$-term $v$ of the form $\mu x. C( \vec w )$, 
a subterm $u$ of $v$ is a \emph{bisimilar subterm} if and only if $v\interp{u}^\emptyset \vsim v$.
We say $v$ is \emph{normal} if and only if all of its bisimilar subterms are $x$,
and all of its strict subterms that do not contain $v$ are also normal.
For instance, the $\mu$-term $v$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is not normal, 
since 
$v\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( v\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( v\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( v\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$,
which is syntactically equivalent to $v$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.

@@

%For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
%(which we will call the \emph{normal form for $v$})
%that is observationally equivalent to $v$.
%We construct $\nf{v}$ by replacing all of the bisimilar subvalues of $v$ with $x$,
%and by replacing all of the non-normal closed subvalues of $v$ with their corresponding normal forms.
%For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.

@@

\(
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  c \text{ normal}  
  \quad
  c \not\sim c' \text{ for all } c' \in \vrange(\Val)
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto c \}
}
\)

Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.

Let $\Val^\ast$ be the result of applying the above rule to completion on $\Val$.
We first argue that $FV( \Val^\ast ) = \emptyset$.
Since \rn{Split} does not apply to $\Fc$,
all equivalence classes of $\Fc$ finite datatype type contain constructor terms, 
and thus $FV( \Val )$ contains no variables of finite datatype type.
For each equivalence classes of $\Fc$ with finite co-datatype $\tau$, 
we know $\tau$ must have cardinality one, due to \rem{reference}.
Since \rn{Singleton} does not apply to $\Fc$, 
there is only one equivalence class of type $\tau$, call it $\ec{x}$.
If $\Val \ec{x} = v_{\ec{x}}$, then we may apply the above rule for the (single) value $c$ of type $\tau$, 
for which the second premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.
Thus, all variables in $FV( \Val )$ have infinite type.
